---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
  )
```

```{r, echo=FALSE, results='asis'}
envreportutils::devex_badge("exploration")
```

[![Travis-CI Build Status](https://travis-ci.org/bcgov/wqbc.svg?branch=master)](https://travis-ci.org/bcgov/wqbc)

---

# wqbc
## Water Quality Thresholds and Index Calculations for British Columbia

The `wqbc` R package calculates water quality thresholds and water quality indices
and plots water quality indices spatially and temporally for
British Columbia.

This package was written for the B.C. Ministry of Environment by [Poisson Consulting](http://www.poissonconsulting.ca/). Ministry of Environment staff maintain the package.

### Usage

For information on use, please see the [vignette](https://htmlpreview.github.com/?https://github.com/bcgov/wqbc/master/inst/doc/wqbc.html). In your R session, you can type `vignette("wqbc")` to see the vignette.

### Install

To install the latest version of wqbc:

```{r, eval = FALSE}
# install the devtools package if it's not already installed
# install.packages("devtools")
library(devtools)
install_github("bcgov/wqbc", build_vignettes = TRUE)
# Using `build_vignettes = TRUE` will slow down the install, but is necessary if 
# you want to read the vignette, which is recommended
library(wqbc)
```

### Project Status

This package is under development.

### Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/wqbc/issues/).

### How to Contribute

If you would like to contribute to the package, please see our 
[CONTRIBUTING](CONTRIBUTING.md) guidelines.

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.

### License

    Copyright 2015 Province of British Columbia

    Licensed under the Apache License, Version 2.0 (the "License");
    you may not use this file except in compliance with the License.
    You may obtain a copy of the License at 

       http://www.apache.org/licenses/LICENSE-2.0

    Unless required by applicable law or agreed to in writing, software
    distributed under the License is distributed on an "AS IS" BASIS,
    WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
    See the License for the specific language governing permissions and
    limitations under the License.
    
This repository is maintained by [Environmental Reporting BC](http://www2.gov.bc.ca/gov/content?id=FF80E0B985F245CEA62808414D78C41B). Click [here](https://github.com/bcgov/EnvReportBC-RepoList) for a complete list of our repositories on GitHub.    
